\documentstyle[11pt]{letter}
\begin{document}

The computer program Big Deal uses a one-to-one correspondence
between numbers and bridge-deals, i.e. each possible bridge-deal
receives its own 'personal identification' number. It uses a
random number generator to produce such an identification number,
which is then converted into a bridge-deal. The conversion from
a given number to its corresponding bridge-deal is handled by the
procedure 'code-to-hand' in the program. This note will explain
the mechanism of this procedure.

Let $G$ denote the total number of possible bridge-deals (one
can show that $G = {52\choose 13}{39\choose 13}{26\choose 13}$).
There are of course numerous ways to establish a one-to-one
correspondence between all possible bridge-deals and the numbers
0 to $G-1$, so we will start explaining the particular choice
that was made for Big Deal.

First, numbers are assigned to the cards. This was done as follows:
Ace of Spades gets number~1, King of Spades is number 2, and so on.
After the Spades are numbered, it is the turn of the Hearts, then
Diamonds and finally Clubs. Two of Clubs is therefore the last card
in the sequence and it has number 52.

Let us now look at a possible hand for North. Reading the card numbers
from low to high, we see that this hand
corresponds to a strictly increasing sequence of 13 numbers,
where the numbers are taken from the set $\{1,2,\ldots,52\}$.
For mathematical convenience, we wish to make a slight
modification to this observation: if, in this strictly
increasing sequence, we subtract $i$ from the $i$-th number,
we obtain a non-decreasing sequence of 13 numbers, where
the numbers are taken from the set $\{0,1,\ldots,39\}$. We
conclude that there is a one-to-one correspondence between
the hands for North and the non-decreasing number sequences of
length 13, where the numbers are taken from $\{0,1,\ldots,39\}$.

Now, knowing North's hand, we turn to East and we wonder if we
can find a one-to-one correspondence between the possible hands
for East and number sequences, similar to the way we did for
North. Of course, one can still argue that East's hand can be
recovered from a non-decreasing number sequences of length 13,
with numbers taken from $\{0,1,\ldots,39\}$. This is true, but
we do not obtain a one-to-one correspondence in this way. To
this end, number the remaining cards (the cards that are not in
North's hand) from 1 to 39. We see that the cards in East's hand
correspond to a strictly increasing number sequence of length 13,
where the numbers are taken from the set $\{1,2,\ldots,39\}$.
Again, we prefer to have a slightly modified representation.
In this strictly increasing sequence we subtract $i$ from the
$i$-th number, thus obtaining a non-decreasing sequence of 13
numbers, where the numbers are taken from the set $\{0,1,\ldots,26\}$.
We conclude that there is a one-to-one correspondence between
the hands for North and the non-decreasing number sequences of
length 13, where the numbers are taken from $\{0,1,\ldots,26\}$.

Knowing both North's and East's hand, there is a one-to-one
correspondence between the possible hands for South and the
non-decreasing number sequences of length 13, where the
numbers are taken from $\{0,1,\ldots,13\}$. Finally, if the
hands of North, East and South are known, then also
West's hand is known, so no coding is needed for West.

So, we can represent any bridge-deal in a unique way by a
non-decreasing number sequence of length 13 with numbers
from $\{0,\ldots,39\}$, followed by a non-decreasing sequence
of length 13 with numbers from $\{0,\ldots,26\}$, terminated
by a non-decreasing sequence of length 13 with numbers from
$\{0,\ldots,13\}$. Let us refer to such a sequence as a
{\it bridge-sequence}. The representation of bridge-deals
by bridge-sequences suggests the following ordering of the
possible bridge-deals. The bridge-deal corresponding to the
sequence consisting of 39 zeros is first in this ordering.
The sequence consisting of 13 times 39, followed by 13 times
26, and terminated by 13 times 13 is last in this ordering.
For any bridge-sequence, unequal to the last one, say
$a_0,\ldots,a_{12},b_0,\ldots,b_{12},c_0,\ldots,c_{12}$,
we construct its {\it successor} as follows. Search for the
right-most number in the sequence that is not at its maximum
(i.e. less than 39 for one of the numbers $a_i$, less than
26 for one of the number $b_i$ or less than 13 for one of
the numbers $c_i$). Let us denote this element of the sequence
by $x_\ell$, where $x$ stands for one the letters $a,b,c$ and
$\ell\in \{0,\ldots,12\}$. The number $x_\ell$ is increased
by~1. The numbers left from it remain unchanged, and the numbers
right from it are set to the lowest possible value that still
results in a valid bridge-sequence: the number $y_k$ is set
to $x_\ell + 1$ if the letter $y$ (seen as a variable taken from
the set $\{a,b,c\}$) equals the letter $x$, and if $k>\ell$. It
is set to zero if the letter $y$ is lexicographically greater than
$x$. This works much like a kilometer counter. The differences
with a normal kilometer counter are only to ensure that the
successor of any bridge-sequence is again a valid bridge-sequence.
Starting at the zero-sequence and incrementing in the way prescribed
above, our kilometer counter will reach any given bridge-sequence.
It is now clear how we intend to assign numbers to the bridge-sequences.
The zero-sequence is assigned~0, its successor is assigned~1, etc. Let
us call this number the G\"odel number of a bridge-sequence.

It turns out that an explicit formula can be given for
the G\"odel number of a bridge-sequence. For the bridge-sequence
$s := (a_0,\ldots,a_{12},b_0,\ldots,b_{12},c_0,\ldots,c_{12})$,
define
$$ \displaystyle
G(s) =  {39 \choose 13} {26 \choose 13} G^a(s)
 +  {26 \choose 13} G^b(s)
 + G^c(s), $$
where
$$G^a(s) =
\displaystyle \sum_{j=0}^{12} ({{52-a_{j-1} - j} \choose {13-j}} - {{52-a_j-j} \choose
{13-j}}), $$
$$G^b(s) =
\displaystyle \sum_{j=0}^{12} ({{39-b_{j-1} - j} \choose {13-j}} - {{39-b_j-j} \choose
{13-j}}), $$ and
$$G^c(s) =
\displaystyle \sum_{j=0}^{12} ({{26-c_{j-1} - j} \choose {13-j}} - {{26-c_j-j} \choose
{13-j}}). $$
In these formulae we should read $a_{-1} = 0$, $b_{-1} = 0$ and $c_{-1} = 0$.
We claim that $G(s)$ is the G\"odel number of $s$. This claim is trivially true
for the zero-sequence. Now it suffices to prove that $G(s^\prime) = G(s) + 1$
whenever $s^\prime$ is the successor of $s$.

Let $x_\ell$ denote the right-most number in the sequence $s$ that is not maximal.
Let us assume for the moment that the variable letter $x$ stands for the letter
$a$. Then $G^b(s^\prime) = G^c(s^\prime) = 0$, since the last 26 numbers in the
$s^\prime$-sequence are 0. Further, we have $G^b(s) = {39\choose 13} - 1$
and $G^c(s) = {26\choose 13} - 1$, since the last 26 numbers in the
$s$-sequence are all maximal. Finally, note that
$$\begin{array}{lll}
G^a(s^\prime) - G^a(s) & = &
\displaystyle - {52-a_{\ell-1}-\ell \choose 13-\ell}
+ {52-a_\ell-\ell \choose 13-\ell} \\
& & \displaystyle - {52-a_{\ell-2}-\ell + 1 \choose 13-\ell + 1}
+ {52-a_{\ell-1}-\ell + 1 \choose 13-\ell + 1} \\
& & \displaystyle + {52-a_{\ell-2}-\ell + 1 \choose 13-\ell + 1}
- {52-a_{\ell-1}-\ell \choose 13-\ell + 1} \\
& = & 1. \end{array} $$
We conclude that indeed
$$G(s^\prime) = G(s) + {39\choose 13}{26\choose 13} - {26\choose 13}({39\choose 13} - 1)
- ({26\choose 13} - 1) = G(s) + 1.$$
The proof that $G(s^\prime) = G(s) + 1$ is similar when the right-most element
that is not maximal is positioned in the $b$-block or the $c$-block, instead
of the $a$-block.

The procedure 'code-to-hand' in the program Big Deal converts a given G\"odel
number to its associated bridge-sequence. The pseudo-code for computing the
first 13 elements in the bridge-sequence is given below. \vspace*{10mm}

{\bf Input:} A (G\"odel) number $g\in\{0,1,\ldots,G-1\}$.
\begin{tabbing}
bb\=bb\=bb\=bb\=bb \kill \\
$a_{-1} := 0;$ \\
$g_{-1} := g;$ \\
{\bf for} $j := 0$ {\bf to} 12 {\bf do} \\
{\bf begin} \\
\>
$a_j :=\displaystyle  \max\{a \mid
{39\choose 13} {26\choose 13} ({52-a_{j-1}-j \choose 13-j} - {52-a-j \choose
13-j}) \leq g_{j-1} \};$ \\
\>
$x_j := \displaystyle
{39\choose 13} {26\choose 13} ({52-a_{j-1}-j \choose 13-j} - {52-a_j-j \choose
13-j});$ \\
\> $g_j := g_{j-1} - x_j;$\\
{\bf end};
\end{tabbing}

In order to show that this code correctly computes the first
13 elements of the bridge-sequence associated with $g$, we
will prove by induction that $$0\leq a_{j-1} \leq a_j \leq 39$$
for all $j\in\{0,\ldots,12\}$. We will also prove that
$$g_j + \sum_{i=0}^j x_i = g$$ for all $j\in \{-1,0,\ldots,12\}$,
and that $$\displaystyle 0\leq g_j < {39\choose 13}{26\choose 13}
({52-a_j-j\choose 13-j} - {52-a_j-1-j\choose 13-j})$$
for all $j\in \{-1,0,\ldots,12\}$.

For $j=-1$, the second claim is trivially true.
For $j=-1$, the third claim states that
$g_{-1} < {39\choose 13}{26\choose 13} ({53\choose 14} - {52\choose 14})$.
Note that ${53\choose 14} - {52\choose 14} = {52\choose 13}$, hence this
claim is equivalent to $g_{-1} < G$, which is also trivially true.

Now, let $j\geq 0$ and assume that the second and third claim are true for
$j-1$. We will show that all three claims are true for $j$.
Note that the inequality
$$ {39\choose 13} {26\choose 13} ({52-a_{j-1} - j
\choose 13-j} - {52-a-j) \choose 13-j}) \leq g^{j-1}$$ is satisfied for
$a = a_{j-1}$, hence $a_j\geq a_{j-1}$. From the fact that the third
claim is true for $j-1$, it follows that this equality is violated
for all $a>39$. Hence, the number $a_j$ is well-defined, and it satisfies
$0\leq a_{j-1} \leq a_j \leq 39$, i.e. the first claim is true for $j$.

The proof that the second claim is true for $j$ follows trivially
by combining $g_j = g_{j-1} - x_j$ and $g_{j-1} + \sum_{i=0}^{j-1} x_i = g$.

It follows from the definition of $a_j$ that
$${39\choose 13}{26\choose 13} ({52-a_{j-1}-j\choose 13-j} -
{52-a_j-1-j)\choose 13-j}) > g_{j-1}.$$
Using this inequality we obtain
$$g_j = g_{j-1} - x_j <
{39\choose 13}{26\choose 13}
({52-a_j-j\choose 13-j} - {52-a_j-1-j\choose 13-j}),$$
and we see that also the third claim is true for $j$.

The computation of the next 13 elements in the bridge-sequence can be done by
similar code. The input-number for this second part is the value $g_{12}$
computed in the first part. Similarly, one can show that this will produce a
non-decreasing sequence $b_0,\ldots,b_{12}$ with elements from
$\{0,\ldots,26\}$, Numbers $g_{13},\ldots,g_{25}$ and $x_{13},\ldots,x_{25}$
will be generated satisfying $$\displaystyle 0\leq g_{j+13} < {26\choose 13}
({39-b_j-j\choose 13-j} - {39-b_j-1-j\choose 13-j})$$ for all $j\in
\{-1,0,\ldots,12\}$, and $$\displaystyle g_j + \sum_{i=0}^j x_i = g$$ for all
$j\in \{-1,0,\ldots,25\}$.

Finally, in the third part a non-decreasing sequence $c_0,\ldots,c_{12}$
with elements from $\{0,\ldots,13\}$ is computed together with
numbers $g_{26},\ldots,g_{38}$ and $x_{26},\ldots,x_{38}$ satisfying
$$\displaystyle 0\leq g_{j+26} < ({26-c_j-j\choose 13-j} - {26-c_j-1-j\choose
13-j})$$ for all $j\in \{-1,0,\ldots,12\}$, and
$$\displaystyle g_j + \sum_{i=0}^j x_i = g$$ for
all $j\in \{-1,0,\ldots,38\}$.

The outcome of the whole procedure is therefore a valid bridge-sequence,
say $s$. The G\"odel number of this bridge-sequence is easily seen to satisfy
$$\displaystyle G(s) = \sum_{i=0}^{38} x_i.$$
Applying the result $$\displaystyle g_j + \sum_{i=0}^j x_i = g$$ for
$j = 38$, we obtain $$G(s) + g_{38} = g.$$ Applying the result
$$\displaystyle 0\leq g_{j+26} < ({26-c_j-j\choose 13-j} - {26-c_j-1-j\choose
13-j})$$ for $j = 12$, we obtain $$\displaystyle 0\leq g_{38} < {14-c_{12}\choose 1}
- {13-c_{12}\choose 1} = 1,$$ hence $g_{38} = 0$. Therefore $G(s) = g$, and we see
that we have actually computed the (unique) bridge-sequence with G\"odel number $g$.

\end{document}
